---
title: "Beta Regression in R"
author:
  - name: Francisco Cribari-Neto
    affiliation: Universidade Federal de Pernambuco
    address: Brazil
    orcid: 0000-0002-5909-6698
    email: cribari@ufpe.br
    url: https://www.de.ufpe.br/~cribari/
  - name: Achim Zeileis
    affiliation: Universität Innsbruck
    address: Austria
    orcid: 0000-0003-0918-3766
    email: Achim.Zeileis@R-project.org
    url: https://www.zeileis.org/
format: 
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
abstract: >
  This introduction to the R package **betareg** is a (slightly)
  modified version of @betareg:Cribari-Neto+Zeileis:2010, published in the
  _Journal of Statistical Software_. A follow-up paper with various
  extensions is @betareg:Gruen+Kosmidis+Zeileis:2012 -- a slightly modified
  version of which is also provided within the package as
  `vignette("betareg-ext", package = "betareg")`\  

  The class of beta regression models is commonly used by 
  practitioners to model variables that assume values in the standard 
  unit interval $(0, 1)$. It is based on the assumption that the 
  dependent variable is beta-distributed and that its mean is related to a set
  of regressors through a linear predictor with unknown coefficients
  and a link function. The model also includes a precision parameter
  which may be constant or depend on a (potentially different)
  set of regressors through a link function as well. This approach naturally
  incorporates features such as heteroskedasticity or skewness
  which are commonly observed in data taking values in the standard
  unit interval, such as rates or proportions. This paper describes
  the **betareg** package which provides the class of beta regressions
  in the R system for statistical computing. The
  underlying theory is briefly outlined, the implementation discussed
  and illustrated in various replication exercises.
bibliography: betareg.bib
vignette: >
  %\VignetteIndexEntry{Beta Regression in R}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{stats,betareg,car,lmtest,sandwich,strucchange}
  %\VignetteKeywords{beta regression, rates, proportions, R}
  %\VignettePackage{betareg}
---

```{r preliminaries, include = FALSE}
library("betareg")
data("GasolineYield", package = "betareg")
data("FoodExpenditure", package = "betareg")
gy_loglog <- betareg(yield ~ batch + temp, data = GasolineYield,
  link = "loglog")
fe_beta2 <- betareg(I(food/income) ~ income + persons | persons,
  data = FoodExpenditure)

knitr::opts_chunk$set(
  engine = "R",
  collapse = TRUE,
  comment = "##",
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
options(width = 70, prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, digits = 5)
```


## Introduction {#sec-intro}

How should one perform a regression analysis in which the dependent variable
(or response variable), $y$, assumes values in the standard unit interval $(0,1)$? The usual
practice used to be to transform the data so that the transformed response, say
$\tilde y$, assumes values in the real line and then apply a standard linear
regression analysis. A commonly used transformation is the logit, $\tilde y = 
\log(y/(1-y))$. This approach, nonetheless, has shortcomings. First, the
regression parameters are interpretable in terms of the mean of $\tilde y$, and
not in terms of the mean of $y$ (given Jensen's inequality). Second, regressions
involving data from the unit interval such as rates and proportions are
typically heteroskedastic: they display more
variation around the mean and less variation as we approach the lower and upper
limits of the standard unit interval. Finally, the distributions of rates and
proportions are typically asymmetric, and thus Gaussian-based approximations for
interval estimation and hypothesis testing can be quite inaccurate in small
samples. @betareg:Ferrari+Cribari-Neto:2004 proposed a regression model
for continuous variates that assume values in the standard unit interval, e.g.,
rates, proportions, or concentration indices. Since the model is
based on the assumption that the response is beta-distributed, they called their
model _the beta regression model_. In their model, the regression
parameters are interpretable in terms of the mean of $y$ (the variable of
interest) and the model is naturally heteroskedastic and easily
accomodates asymmetries. A variant of the beta regression model that allows for
nonlinearities and variable dispersion was proposed by
@betareg:Simas+Barreto-Souza+Rocha:2010. In particular, in this more
general model, the parameter accounting for the precision of the data is not
assumed to be constant across observations but it is allowed to vary, leading to the
_variable dispersion beta regression model_. 

The chief motivation for the beta regression model lies in the flexibility
delivered by the assumed beta law. The beta density can assume a number of
different shapes depending on the combination of parameter values, including
left- and right-skewed or the flat shape of the uniform density (which is a
special case of the more general beta density). This is illustrated in
@fig-beta-distributions which depicts several different beta densities.
Following @betareg:Ferrari+Cribari-Neto:2004, the densities are
parameterized in terms of the mean $\mu$ and the precision parameter $\phi$;
all details are explained in the next section. The evident flexiblity makes
the beta distribution an attractive candidate for data-driven statistical modeling.

```{r beta-distributions}
#| echo: false
#| fig-width: 8.5
#| fig-height: 4.5
#| out-width: 100%
#| label: fig-beta-distributions
#| fig-cap: "Probability density functions for beta distributions with varying parameters $\\mu = 0.10, 0.25, 0.50, 0.75, 0.90$ and $\\phi = 5$ (left) and $\\phi = 100$ (right)."
par(mfrow = c(1, 2), mar = c(4.1, 4.1, 4.1, 0.1))
dbeta2 <- function(x, mu, phi = 1) dbeta(x, mu * phi, (1 - mu) * phi)
x <- seq(from = 0.01, to = 0.99, length = 200)
xx <- cbind(x, x, x, x, x)

yy <- cbind(
  dbeta2(x, 0.10, 5),
  dbeta2(x, 0.25, 5),
  dbeta2(x, 0.50, 5),
  dbeta2(x, 0.75, 5),
  dbeta2(x, 0.90, 5)
)
matplot(xx, yy, type = "l", xlab = "y", ylab = "Density", main = expression(phi == 5),
  lty = 1, col = "black", ylim = c(0, 15))
text(0.05, 12  , "0.10")
text(0.95, 12  , "0.90")
text(0.22,  2.8, "0.25")
text(0.78,  2.8, "0.75")
text(0.50,  2.3, "0.50")

yy <- cbind(
  dbeta2(x, 0.10, 100),
  dbeta2(x, 0.25, 100),
  dbeta2(x, 0.50, 100),
  dbeta2(x, 0.75, 100),
  dbeta2(x, 0.90, 100)
)
matplot(xx, yy, type = "l", xlab = "y", ylab = "", main = expression(phi == 100),
  lty = 1, col = "black", ylim = c(0, 15))
text(0.10, 14.5, "0.10")
text(0.90, 14.5, "0.90")
text(0.25,  9.8, "0.25")
text(0.75,  9.8, "0.75")
text(0.50,  8.6, "0.50")
```

The idea underlying beta regression models dates back to earlier approaches such as
@betareg:Williams:1982 or @betareg:Prentice:1986. The initial
motivation was to model binomial random variables with extra variation. The
model postulated for the (discrete) variate of interest included a more flexible
variation structure determined by independent beta-distributed variables which
are related to a set of independent variables through a regression structure.
However, unlike the more recent literature, the main focus
was to model binomial random variables. Our interest in what follows will be 
more closely related to the recent literature, i.e., modeling continous random variables that
assume values in $(0,1)$, such as rates, proportions, and concentration or
inequality indices (e.g., Gini).

In this paper, we describe the **betareg** package which can
be used to perform inference in both fixed and variable dispersion beta
regressions. The package is implemented in the R system
for statistical computing [@betareg:R:2009] and available
from the Comprehensive R Archive Network (CRAN) at
<https://CRAN.R-project.org/package=betareg>. The initial version of the
package was written by @betareg:Simas+Rocha:2006 up to version 1.2
which was orphaned and archived on CRAN in mid-2009. Starting from version 2.0-0,
Achim Zeileis took over maintenance after rewriting/extending the package's
functionality.

The paper unfolds as follows:
@sec-model outlines the theory underlying the beta regression model
before @sec-implementation describes its implementation in R.
@sec-illustrations and @sec-replications provide various
empirical applications: The former focuses on illustrating various aspects
of beta regressions in practice while the latter provides further replications
of previously published empirical research. Finally, @sec-conclusion
contains concluding remarks and directions for future research and 
implementation. 


## Beta regression {#sec-model} 

The class of beta regression models, as introduced by
@betareg:Ferrari+Cribari-Neto:2004, is useful for modeling continuous
variables $y$ that assume  values in the open standard unit interval $(0,1)$.
Note that if the variable takes on values in $(a, b)$ (with $a < b$ known) one
can model $(y - a)/(b - a)$. Furthermore, if $y$ also assumes the extremes $0$ and $1$, 
a useful transformation in practice is $(y \cdot (n - 1) + 0.5) / n$
where $n$ is the sample size [@betareg:Smithson+Verkuilen:2006].

The beta regression model is based on an alternative parameterization of the beta density in
terms of the  variate mean and a precision parameter. The beta density is
usually expressed as 

$$
f(y;p,q) = \frac{\Gamma(p+q)}{\Gamma(p)\Gamma(q)}y^{p-1}(1-y)^{q-1}, \quad 0<y<1, 
$$

where $p,q >0$ and $\Gamma(\cdot)$ is the gamma function.^[A beta regression model
based on this parameterization was proposed by @betareg:Vasconcellos+Cribari-Neto:2005.
We shall, however, focus on the parameterization indexed by the mean and a precision
parameter.]
@betareg:Ferrari+Cribari-Neto:2004 proposed a different parameterization
by setting $\mu = p/(p+q)$ and $\phi = p+q$:

$$
f(y;\mu,\phi) = \frac{\Gamma(\phi)}{\Gamma(\mu\phi)\Gamma((1-\mu)\phi)}y^{\mu\phi-1}(1-y)^{(1-\mu)\phi-1}, \quad 0<y<1,
$$ {#eq-density0}

with $0<\mu<1$ and $\phi>0$.
We write $y \,\sim\, \mathcal{B}(\mu, \phi)$. Here, $\text{E}(y) = \mu$ and $\text{Var}(y) = \mu(1-\mu)/(1+\phi)$. The
parameter $\phi$ is known as the precision parameter since, for fixed $\mu$, the larger $\phi$ the smaller the variance of $y$; $\phi^{-1}$ is a dispersion parameter. 

Let $y_1,\ldots,y_n$ be a random sample such that $y_i \sim 
{\mathcal{B}}(\mu_i,\phi)$, $i=1,\ldots,n$. The beta regression model is 
defined as 

$$
g(\mu_i) = x_{i}^\top \beta = \eta_i,
$$

where $\beta=(\beta_1,\ldots,\beta_k)^\top$ is a $k \times 1$ vector of unknown 
regression parameters ($k<n$), $x_i = (x_{i1},\ldots,x_{ik})^\top$ is the vector 
of $k$ regressors (or independent variables or covariates) and $\eta_i$ is a linear 
predictor (i.e., $\eta_i = \beta_1 x_{i1} + \cdots + \beta_k x_{ik}$; usually 
$x_{i1}=1$ for all $i$ so that the model has an intercept).  
Here, $g(\cdot): (0,1) \mapsto
\mathrm{I\! R}$ is a link function, which is strictly increasing and twice
differentiable. The main motivation for using a link function in the regression structure is twofold. First,
both sides of the regression equation assume values in the real line when a link function is applied to
$\mu_i$. Second, there is an added flexibility since the practitioner can choose the function that yields the best fit. 
Some useful link functions are: logit $g(\mu) = \log(\mu/(1-\mu))$;
probit $g(\mu) = \Phi^{-1}(\mu)$, where $\Phi(\cdot)$ is the standard normal
distribution function; complementary log-log $g(\mu) = \log\{-\log(1-\mu)\}$;
log-log $g(\mu) = -\log\{-\log(\mu)\}$; and Cauchy $g(\mu) = \tan\{\pi(\mu - 0.5)\}$.
Note that the variance of $y$ is a function of $\mu$ which renders the regression
model based on this parameterization naturally heteroskedastic. In particular, 

$$
   \text{Var}(y_i) = \frac{\mu_i(1-\mu_i)}{1+\phi}
             = \frac{g^{-1}(x_i^{\top}\beta)[1-g^{-1}(x_i^{\top}\beta)]}{1+\phi}.
$$ {#eq-variance}
  
The log-likelihood function is $\ell(\beta,\phi)=\sum_{i=1}^n\ell_i(\mu_i,\phi)$, where 

$$
\begin{eqnarray}
  \ell_i(\mu_i,\phi) & = & \log \Gamma(\phi)-\log\Gamma(\mu_i\phi) - \log \Gamma((1-\mu_i)\phi) +(\mu_i\phi-1)\log y_i \\
                     &   & + \{(1-\mu_i)\phi-1\}\log(1-y_i).
\end{eqnarray}
$$ {#eq-loglik}

Notice that $\mu_i=g^{-1}(x_i^{\top}\beta)$ is a function of $\beta$, the vector
of regression parameters. Parameter estimation is performed by maximum likelihood (ML).

An extension of the beta regression model above which was employed by
@betareg:Smithson+Verkuilen:2006 and formally introduced (along
with further extensions) by @betareg:Simas+Barreto-Souza+Rocha:2010
is the variable dispersion beta regression model. In this model the
precision parameter is not constant for all observations but instead
modeled in a similar fashion as the mean parameter. More specifically,
$y_i \, \sim \, {\mathcal B}(\mu_i, \phi_i)$ independently, $i=1,\ldots,n$, and 

$$
\begin{eqnarray}
  g_1(\mu_i)  & = & \eta_{1i} = x_i^\top \beta,  \\
  g_2(\phi_i) & = & \eta_{2i} = z_i^\top \gamma,
\end{eqnarray}
$$ {#eq-meanprecision}

where $\beta=(\beta_1, \ldots, \beta_k)^{\top}$,
$\gamma=(\gamma_1,\ldots,\gamma_h)^{\top}$, $k+h<n$, are the sets of
regression coefficients in the two equations, $\eta_{1i}$ and $\eta_{2i}$ are
the linear predictors, and $x_i$ and $z_i$ are regressor vectors. As before,
both coefficient vectors are estimated by ML, simply replacing $\phi$ by $\phi_i$
in @eq-loglik.

@betareg:Simas+Barreto-Souza+Rocha:2010 further extend the model above
by allowing nonlinear predictors in @eq-meanprecision. Also, they have obtained analytical bias
corrections for the ML estimators of the parameters, thus generalizing the results of
@betareg:Ospina+Cribari-Neto+Vasconcellos:2006, who derived bias
corrections for fixed dispersion beta regressions. However, as these extensions
are not (yet) part of the **betareg** package, we confine ourselves to these
short references and do not provide detailed formulas.

Various types of residuals are available for beta regression models. The raw
response residuals $y_i - \hat \mu_i$ are typically not used due to the heteroskedasticity
inherent in the model (see @eq-variance). Hence, a natural
alternative are _Pearson residuals_ which @betareg:Ferrari+Cribari-Neto:2004
call _standardized ordinary residuals_ and define as

$$
  r_{\mathrm{P}, i} = \frac{y_i - \hat{\mu}_i}{\sqrt{\widehat{\text{Var}}(y_i)}},
$$ {#eq-pearson}

where $\widehat{\text{Var}}(y_i) = \hat{\mu}_i(1-\hat{\mu}_i)/
(1+\hat{\phi_i})$, $\hat{\mu}_i = g_1^{-1}(x_i^\top\,\hat{\beta})$,
and $\hat{\phi}_i = g_2^{-1}(z_i^\top\,\hat{\gamma})$. Similarly, deviance residuals
can be defined in the standard way via signed contributions to the excess likelihood.
Further residuals were proposed by @betareg:Espinheira+Ferrari+Cribari-Neto:2008a,
in particular one residual with better properties that they named
_standardized weighted residual 2_:

$$
  r_{\mathrm{sw2}, i} = \frac{y^*_i - {\hat{{\mu}}^*}_i}{\sqrt{\hat{v}_i(1 - h_{ii})}},
$$ {#eq-sweighted2}

where $y_i^* = \log\{ y_i / (1-y_i)\}$ and  $\mu_i^* = \psi(\mu_i\phi)- \psi((1-\mu_i)\phi)$,
$\psi(\cdot)$ denoting the digamma function. Standardization is then by
$v_i = \left\{ \psi'(\mu_i\phi) + \psi'((1-\mu_i)\phi)\right\}$
and $h_{ii}$, the $i$th diagonal element of the hat matrix
\citep[for details see][]{betareg:Ferrari+Cribari-Neto:2004,betareg:Espinheira+Ferrari+Cribari-Neto:2008a}.
Hats denote evaluation at the ML estimates.

It is noteworthy that the beta regression model described above was developed to 
allow practitioners to model continuous variates that assume values in the 
unit interval such as rates, proportions, and concentration or inequality
indices (e.g., Gini). However, the data types that can be modeled using beta regressions
also encompass proportions of "successes" from a number of trials, if the
number of trials is large enough to justify a continuous model. In this
case, beta regression is similar to a binomial generalized linear model (GLM)
but provides some more flexibility -- in particular when the trials are not
independent and the standard binomial model might be too strict. 
In such a situation, the fixed dispersion beta regression is similar to
the quasi-binomial model [@betareg:McCullagh+Nelder:1989] but fully
parametric. Furthermore, it can be naturally extended to variable dispersions.


## Implementation in R {#sec-implementation}

To turn the conceptual model from the previous section into computational tools
in R, it helps to emphasize some properties of the model: It is a standard
maximum likelihood (ML) task for which there is no closed-form solution but numerical
optimization is required. Furthermore, the model shares some properties (such as
linear predictor, link function, dispersion parameter) with
generalized linear models [GLMS, @betareg:McCullagh+Nelder:1989], but it
is not a special case of this framework (not even for fixed dispersion). 
There are various
models with implementations in R that have similar features -- here, we
specifically reuse some of the ideas employed for generalized count data regression
by @betareg:Zeileis+Kleiber+Jackman:2008.

The main model-fitting function in **betareg** is `betareg()` which takes
a fairly standard approach for implementing ML regression models in R:
`formula` plus `data` is used for model and data
specification, then the likelihood and corresponding gradient (or estimating
function) is set up, `optim()` is called for maximizing the likelihood, and finally
an object of S3 class `"betareg"` is returned for which a large set of methods to standard
generics is available. The workhorse function is `betareg.fit()` which provides
the core computations without `formula`-related data pre- and post-processing.
**Update:** Recently, `betareg()` has been extended to optionally include an
additional Fisher scoring iteration after the `optim()` optimization in order to
improve the ML result (or apply a bias correction or reduction).

The model-fitting function `betareg()` and its associated class are designed to
be as similar as possible to the standard `glm()` function [@betareg:R:2009]
for fitting GLMs. An important difference is that there are potentially two equations
for mean and precision (@eq-meanprecision), and consequently two regressor matrices,
two linear predictors, two sets of coefficients, etc. In this respect, the design
of `betareg()` is similar to the functions described by
@betareg:Zeileis+Kleiber+Jackman:2008 for fitting zero-inflation and hurdle
models which also have two model components. The arguments of `betareg()` are

```{r, eval=FALSE}
betareg(formula, data, subset, na.action, weights, offset,
  link = "logit", link.phi = NULL, control = betareg.control(...),
  model = TRUE, y = TRUE, x = FALSE, ...)
```

where the first line contains the standard model-frame specifications
[see @betareg:Chambers+Hastie:1992],
the second line has the arguments specific to beta regression models and
the arguments in the last line control some components of the return value.

| Function              | Description                                                                   |
|:----------------------|:------------------------------------------------------------------------------|
| `print()`             | Simple printed display with coefficient estimates |
| `summary()`           | Standard regression output (coefficient estimates, standard errors, partial Wald tests); returns an object of class `"summary.betareg"` containing the relevant summary statistics (which has a `print()` method) |
| `coef()`              | Extract coefficients of model (full, mean, or precision components), a single vector of all coefficients by default |
| `vcov()`              | Associated covariance matrix (with matching names) |
| `predict()`           | Predictions (of means $\mu_i$, linear predictors $\eta_{1i}$, precision parameter $\phi_i$, or variances $\mu_i (1 - \mu_i) / (1 + \phi_i)$) for new data |
| `fitted()`            | Fitted means for observed data |
| `residuals()`         | Extract residuals (deviance, Pearson, response, quantile, or different weighted residuals), see @betareg:Espinheira+Ferrari+Cribari-Neto:2008a, defaulting to quantile residuals since version 3.2-0 |
| `estfun()`            | Compute empirical estimating functions (or score functions), evaluated at observed data and estimated parameters, see @betareg:Zeileis:2006a |
| `bread()`             | Extract "bread" matrix for sandwich estimators, see @betareg:Zeileis:2006a |
| `terms()`             | Extract terms of model components |
| `model.matrix()`      | Extract model matrix of model components |
| `model.frame()`       | Extract full original model frame |
| `logLik()`            | Extract fitted log-likelihood |
| `plot()`              | Diagnostic plots of residuals, predictions, leverages etc. |
| `hatvalues()`         | Hat values (diagonal of hat matrix) |
| `cooks.distance()`    | (Approximation of) Cook's distance |
| `gleverage()`         | Compute generalized leverage, see @betareg:Wei+Hu+Fung:1998 and @betareg:Rocha+Simas:2010 |
| `coeftest()`          | Partial Wald tests of coefficients |
| `waldtest()`          | Wald tests of nested models |
| `linear.hypothesis()` | Wald tests of linear hypotheses |
| `lrtest()`            | Likelihood ratio tests of nested models |
| `AIC()`               | Compute information criteria (AIC, BIC, etc.) |

: Functions and methods for objects of class `"betareg"`. The first 17 functions refer to methods, the last five are generic functions whose default methods work because of the information supplied by the methods above. {#tbl-methods}


If a `formula` of type `y ~ x1 + x2` is supplied, it describes $y_i$ and $x_i$ 
for the mean equation of the beta regression (@eq-meanprecision). In this case a constant
$\phi_i$ is assumed, i.e., $z_i = 1$ and $g_2$ is the identity link, corresponding
to the basic beta regression model as introduced in @betareg:Ferrari+Cribari-Neto:2004.
However, a second set of regressors can be specified by a two-part formula of type
`y ~ x1 + x2 | z1 + z2 + z3` as provided in the **Formula** package
[@betareg:Zeileis+Croissant:2010]. This model has the same mean equation as above but
the regressors $z_i$ in the precision equation (@eq-meanprecision) are taken from
the `~ z1 + z2 + z3` part.
The default link function in this case is the log link $g_2(\cdot) = \log(\cdot)$.
Consequently, `y ~ x1 + x2` and `y ~ x1 + x2 | 1` correspond to equivalent
beta likelihoods but use different parametrizations for $\phi_i$: simply $\phi_i = \gamma_1$
in the former case and $\log(\phi_i) = \gamma_1$ in the latter case. The link for the
$\phi_i$ precision equation can be changed by `link.phi` in both cases where
`"identity"`, `"log"`, and `"sqrt"` are allowed as admissible values.
The default for the $\mu_i$ mean equation is always the logit link but all link
functions for the `binomial` family in `glm()` are allowed as well as the log-log
link: `"logit"`, `"probit"`, `"cloglog"`, `"cauchit"`,
`"log"`, and `"loglog"`.

ML estimation of all parameters employing analytical gradients is carried out
using R's `optim()` with control options set in `betareg.control()`.
All of `optim()`'s methods are available but the default is  
`"BFGS"`, which is typically regarded to be the best-performing method
[@betareg:Mittelhammer+Judge+Miller:2000, Section 8.13] with the most
effective updating formula of all quasi-Newton methods [@betareg:Nocedal+Wright:1999, p. 197].
Starting values can be user-supplied, otherwise the $\beta$ starting values are estimated by 
a regression of $g_1(y_i)$ on $x_i$. The starting values for the $\gamma$ intercept
are chosen as described in @betareg:Ferrari+Cribari-Neto:2004 [p. 805], corresponding
to a constant $\phi_i$ (plus a link transformation, if any). All further
$\gamma$ coefficients (if any) are initially set to zero.
The covariance matrix estimate is derived analytically as in
@betareg:Simas+Barreto-Souza+Rocha:2010. However, by setting `hessian = TRUE`
the numerical Hessian matrix returned by `optim()` can also be obtained.
**Update:** In recent versions of **betareg**, the `optim()` is
still performed but optionally it may be complemented by a subsequent
additional Fisher scoring iteration to improve the result.

The returned fitted-model object of class `"betareg"` is a list similar
to `"glm"` objects. Some of its elements -- such as `coefficients` or
`terms` -- are lists with a mean and precision component,
respectively.

A set of standard extractor functions for fitted model objects is available for
objects of class `"betareg"`, including the usual `summary()` method that
includes partial Wald tests for all coefficients. No `anova()` method is provided,
but the general `coeftest()` and `waldtest()` from
**lmtest** [@betareg:Zeileis+Hothorn:2002], and `linear.hypothesis()`
from **car** [@betareg:Fox+Weisberg:2019] can be used for Wald tests while `lrtest()` from **lmtest**
provides for likelihood-ratio tests of nested models. See @tbl-methods for a list
of all available methods. Most of these are standard in base R, however, methods
to a few less standard generics are also provided. Specifically, there are tools related
to specification testing and computation of sandwich covariance matrices as discussed
by @betareg:Zeileis:2004,betareg:Zeileis:2006a as well as a method to a new generic for computing
generalized leverages [@betareg:Wei+Hu+Fung:1998].


## Beta regression in practice {#sec-illustrations}

To illustrate the usage of **betareg** in practice we replicate
and slightly extend some of the analyses from the original papers
that suggested the methodology. More specifically, we estimate and
compare various flavors of beta regression models for the gasoline
yield data of @betareg:Prater:1956, see @fig-GasolineYield,
and for the household food expenditure data taken from
@betareg:Griffiths+Hill+Judge:1993, see @fig-FoodExpenditure). 
Further pure replication exercises are provided in @sec-replications.

### The basic model: Estimation, inference, diagnostics

#### Prater's gasoline yield data


The basic beta regression model as suggested by @betareg:Ferrari+Cribari-Neto:2004
is illustrated in Section 4 of their paper using two empirical examples.
The first example employs the well-known gasoline yield data taken from
@betareg:Prater:1956. The variable of interest is `yield`,
the proportion of crude oil converted to gasoline after distillation
and fractionation, for which a beta regression model is rather
natural. @betareg:Ferrari+Cribari-Neto:2004 employ two explanatory
variables: `temp`, the temperature (in degrees Fahrenheit) at which all
gasoline has vaporized, and `batch`, a factor indicating ten unique batches
of conditions in the experiments (depending on further variables). The
data, encompassing `r nrow(GasolineYield)` observations, is visualized
in @fig-GasolineYield.

@betareg:Ferrari+Cribari-Neto:2004 start out with a model where
`yield` depends on `batch` and `temp`, employing the standard
logit link. In **betareg**, this can be fitted via

```{r GasolineYield-betareg}
data("GasolineYield", package = "betareg")
gy_logit <- betareg(yield ~ batch + temp, data = GasolineYield)
summary(gy_logit)
```

which replicates their Table 1. The goodness of fit is assessed using
different types of diagnostic displays shown in their Figure 2. This
graphic can be reproduced (in a slightly different order) using the
`plot()` method for `"betareg"` objects, see @fig-GasolineYield-plot.

```{r GasolineYield-visualization}
#| echo: false
#| fig-width: 6
#| fig-height: 5.5
#| out-width: 100%
#| label: fig-GasolineYield
#| fig-cap: "Gasoline yield data from @betareg:Prater:1956: Proportion of crude oil converted to gasoline explained by temperature (in degrees Fahrenheit) at which all gasoline has vaporized and given batch (indicated by gray level). Fitted curves correspond to beta regressions `gy_loglog` with log-log link (solid, red) and `gy_logit` with logit link (dashed, blue). Both curves were evaluated at varying temperature with the intercept for batch 6 (i.e., roughly the average intercept)."
redblue <- hcl(c(0, 260), 90, 40)
plot(yield ~ temp, data = GasolineYield, type = "n",
  ylab = "Proportion of crude oil converted to gasoline",
  xlab = "Temperature at which all gasoline has vaporized",
  main = "Prater's gasoline yield data")
points(yield ~ temp, data = GasolineYield, cex = 1.75, 
  pch = 19, col = rev(gray.colors(10))[as.numeric(batch)])
points(yield ~ temp, data = GasolineYield, cex = 1.75)
legend("topleft", as.character(1:10), title = "Batch",
  col = rev(gray.colors(10)), pch = 19, bty = "n")
legend("topleft", as.character(1:10), title = "Batch", pch = 1, bty = "n")
lines(150:500, predict(gy_logit, 
  newdata = data.frame(temp = 150:500, batch = "6")),
  col = redblue[2], lwd = 2, lty = 2)
lines(150:500, predict(gy_loglog, 
  newdata = data.frame(temp = 150:500, batch = "6")),
  col = redblue[1], lwd = 2)
legend("bottomright", c("log-log", "logit"),
  col = redblue, lty = 1:2, lwd = 2, bty = "n")
```


```{r GasolineYield-plot}
#| fig-width: 8.5
#| fig-height: 10
#| out-width: 100%
#| label: fig-GasolineYield-plot
#| fig-cap: "Diagnostic plots for beta regression model `gy_logit`."
par(mfrow = c(3, 2))
suppressWarnings(RNGversion("3.5.0"))
set.seed(123)
plot(gy_logit, which = 1:4, type = "pearson")
plot(gy_logit, which = 5, type = "deviance", sub.caption = "")
plot(gy_logit, which = 1, type = "deviance", sub.caption = "")
```

As observation 4 corresponds to a large Cook's distance and large
residual, @betareg:Ferrari+Cribari-Neto:2004 decided to refit
the model excluding this observation. While this does not change the
coefficients in the mean model very much, the precision parameter
$\phi$ increases clearly.

```{r GasolineYield-update}
gy_logit4 <- update(gy_logit, subset = -4)
coef(gy_logit, model = "precision")
coef(gy_logit4, model = "precision")
```



#### Household food expenditures

@betareg:Ferrari+Cribari-Neto:2004 also consider a second
example: household food expenditure data for `r nrow(FoodExpenditure)` households
taken from @betareg:Griffiths+Hill+Judge:1993 (Table 15.4).
The dependent variable is `food/income`, the proportion of household
`income` spent on `food`. Two explanatory variables are available:
the previously mentioned household `income` and the number of `persons`
living in the household. All three variables are visualized in @fig-FoodExpenditure.

To start their analysis, @betareg:Ferrari+Cribari-Neto:2004 consider
a simple linear regression model fitted by ordinary least squares (OLS):

```{r FoodExpenditure-lm}
data("FoodExpenditure", package = "betareg")
fe_lm <- lm(I(food/income) ~ income + persons, data = FoodExpenditure)
```

To show that this model exhibits heteroskedasticity, they employ 
the studentized @betareg:Breusch+Pagan:1979 test of @betareg:Koenker:1981
which is available in R in the **lmtest** package
[@betareg:Zeileis+Hothorn:2002].

```{r FoodExpenditure-bptest}
library("lmtest")
bptest(fe_lm)
```

One alternative would be to consider a logit-transformed response in
a traditional OLS regression but this would make the residuals
asymmetric. However, both issues -- heteroskedasticity and skewness --
can be alleviated when a beta regression model with a logit link for
the mean is used.

```{r FoodExpenditure-betareg}
fe_beta <- betareg(I(food/income) ~ income + persons,
  data = FoodExpenditure)
summary(fe_beta)
```

This replicates Table 2 from @betareg:Ferrari+Cribari-Neto:2004.
The predicted means of the linear and the beta regression model, respectively,
are very similar: the proportion of household income spent on food decreases
with the overall income level but increases in the number of persons in the
household (see also @fig-FoodExpenditure).

Below, further extended models will be considered for these data sets and
hence all model comparisons are deferred.

```{r FoodExpenditure-visualization}
#| echo: false
#| fig-width: 6
#| fig-height: 5.5
#| out-width: 100%
#| label: fig-FoodExpenditure
#| fig-cap: "Household food expenditure data from @betareg:Griffiths+Hill+Judge:1993: Proportion of household income spent on food explained by household income and number of persons in household (indicated by gray level). Fitted curves correspond to beta regressions `fe_beta` with fixed dispersion (long-dashed, blue), `fe_beta2` with variable dispersion (solid, red), and the linear regression `fe_lin` (dashed, black). All curves were evaluated at varying income with the intercept for mean number of persons ($ = `r round(mean(FoodExpenditure$persons), digits = 2)`$)."
redblueblack <- hcl(c(0, 260, 0), c(90, 90, 0), c(40, 40, 0))
plot(I(food/income) ~ income, data = FoodExpenditure,
  xlab = "Household income", ylab = "Proportion of food expenditures",
  main = "Food expenditures data", type = "n", ylim = c(0.04, 0.57))
points(I(food/income) ~ income, data = FoodExpenditure, cex = 1.75, pch = 19,
  col = rev(gray.colors(7))[persons])
points(I(food/income) ~ income, data = FoodExpenditure, cex = 1.75)
legend("bottomleft", rev(as.character(sort(unique(FoodExpenditure$persons)))),
  title = "Persons", col = gray.colors(7), pch = 19, bty = "n")
legend("bottomleft", rev(as.character(sort(unique(FoodExpenditure$persons)))),
  title = "Persons", pch = 1, bty = "n")
lines(10:100, predict(fe_lm, 
  newdata = data.frame(income = 10:100, persons = mean(FoodExpenditure$persons))),
  col = redblueblack[3], lwd = 2, lty = 2)
lines(10:100, predict(fe_beta, 
  newdata = data.frame(income = 10:100, persons = mean(FoodExpenditure$persons))),
  col = redblueblack[2], lwd = 2, lty = 5)
lines(10:100, predict(fe_beta2, 
  newdata = data.frame(income = 10:100, persons = mean(FoodExpenditure$persons))),
  col = redblueblack[1], lwd = 2)
legend("topright", c("logit, var. disp.", "logit, fix. disp.", "lm"),
  col = redblueblack, lty = c(1, 5, 2), lwd = 2, bty = "n")
```


### Variable dispersion model {#sec-vardisp}

#### Prater's gasoline yield data

Although the beta model already incorporates naturally a certain
pattern in the variances of the response (see @eq-variance), it might
be necessary to incorporate further regressors to account for heteroskedasticity
as in @eq-meanprecision [@betareg:Simas+Barreto-Souza+Rocha:2010].
For illustration of this approach, the example from Section 3 of the online
supplements to @betareg:Simas+Barreto-Souza+Rocha:2010 is considered.
This investigates Prater's gasoline yield data based on the same
mean equation as above, but now with temperature `temp` as an 
additional regressor for the precision parameter $\phi_i$:

```{r GasolineYield-phireg}
gy_logit2 <- betareg(yield ~ batch + temp | temp, data = GasolineYield)
```

for which `summary(gy_logit2)` yields the MLE column in Table 19
of @betareg:Simas+Barreto-Souza+Rocha:2010. To save space, only
the parameters pertaining to $\phi_i$ are reported here

```{r GasolineYield-phireg-coef, echo=FALSE}
printCoefmat(summary(gy_logit2)$coefficients$precision)
```

which signal a significant improvement by including the `temp` regressor.
Instead of using this Wald test, the models can also be compared by means
of a likelihood-ratio test (see their Table 18) that confirms the results:

```{r GasolineYield-lrtest}
lrtest(gy_logit, gy_logit2)
```

Note that this can also be interpreted as testing the null hypothesis of equidispersion
against a specific alternative of variable dispersion.


#### Household food expenditures

For the household food expenditure data, the Breusch-Pagan test carried out
above illustrated that there is heteroskedasticity that can be captured
by the regressors `income` and `persons`. Closer investigation
reveals that this is mostly due to the number of persons in the household,
also brought out graphically by some of the outliers with high values in
this variable in @fig-FoodExpenditure. Hence, it seems natural
to consider the model employed above with `persons` as an additional
regressor in the precision equation.

```{r FoodExpenditure-betareg2}
fe_beta2 <- betareg(I(food/income) ~ income + persons | persons,
  data = FoodExpenditure)
```

This leads to significant improvements in terms of the likelihood and
the associated BIC.^[In R, the BIC can be computed by means of
`AIC()` when $\log(n)$ is supplied as the penalty term `k`.]

```{r FoodExpenditure-comparison}
lrtest(fe_beta, fe_beta2)
AIC(fe_beta, fe_beta2, k = log(nrow(FoodExpenditure)))
```

Thus, there is evidence for variable dispersion and model `fe_beta2` seems
to be preferable. As visualized in @fig-FoodExpenditure, it
describes a similar relationship between response and explanatory variables
although with a somewhat shrunken `income` slope.


### Selection of different link functions

#### Prater's gasoline yield data

As in binomial GLMs, selection of an appropriate link function 
can greatly improve the model fit [@betareg:McCullagh+Nelder:1989],
especially if extreme proportions (close to $0$ or $1$) have been
observed in the data. To illustrate this problem in beta regressions,
we replicate parts of the analysis in Section 5 of @betareg:Cribari-Neto+Lima:2007.
This reconsiders Prater's gasoline yield data but employs a log-log
link instead of the previously used (default) logit link

```{r GasolineYield-loglog}
gy_loglog <- betareg(yield ~ batch + temp, data = GasolineYield,
  link = "loglog")
```

which clearly improves the pseudo $R^2$ of the model:

```{r GasolineYield-Rsquared}
summary(gy_logit)$pseudo.r.squared
summary(gy_loglog)$pseudo.r.squared
```
Similarly, the AIC^[Note that @betareg:Cribari-Neto+Lima:2007 did not
account for estimation of $\phi$ in their degrees of freedom. Hence, their reported
AICs differ by 2.] (and BIC) of the fitted model is not only
superior to the logit model with fixed dispersion `gy_logit`
but also to the logit model with variable dispersion `gy_logit2` considered
in the previous section.

```{r GasolineYield-AIC}
AIC(gy_logit, gy_logit2, gy_loglog)
```
Moreover, if `temp` were included as a regressor in the precision
equation of `gy_loglog`, it would no longer yield significant improvements.
Thus, improvement of the model fit in the mean equation by adoption of the log-log
link has waived the need for a variable precision equation.

To underline the appropriateness of the log-log specification, 
@betareg:Cribari-Neto+Lima:2007 consider a sequence of diagnostic
tests inspired by the RESET [regression specification error test; @betareg:Ramsey:1969]
in linear regression models.
To check for misspecifications, they consider powers of fitted means or linear predictors
to be included as auxiliary regressors in the mean equation. In well-specified
models, these should not yield significant improvements. For the gasoline yield
model, this can only be obtained for the log-log link while all other link
functions result in significant results indicating misspecification. Below,
this is exemplified for a likelihood-ratio test of squared linear predictors.
Analogous results can be obtained for `type = "response"` or higher powers.

```{r GasolineYield-reset}
lrtest(gy_logit, . ~ . + I(predict(gy_logit, type = "link")^2))
lrtest(gy_loglog, . ~ . + I(predict(gy_loglog, type = "link")^2))
```

The improvement of the model fit can also be brought out graphically by comparing
absolute raw residuals (i.e., $y_i - \hat \mu_i$) from both models as in
@fig-GasolineYield-diagnostics.

```{r GasolineYield-diagnostics}
#| echo: false
#| fig-width: 6
#| fig-height: 5.5
#| out-width: 100%
#| label: fig-GasolineYield-diagnostics
#| fig-cap: "Scatterplot comparing the absolute raw residuals from beta regression modes with log-log link (x-axis) and logit link (y-axis)."
plot(abs(residuals(gy_loglog, type = "response")),
  abs(residuals(gy_logit, type = "response")))
abline(0, 1, lty = 2)
```

This shows that there are a few observations clearly above the diagonal
(where the log-log-link fits better than the logit link) whereas
there are fewer such observations below the diagonal. A different
diagnostic display that is useful in this situation 
[and is employed by @betareg:Cribari-Neto+Lima:2007] is a 
plot of predicted values ($\hat \mu_i$) vs. observed values ($y_i$)
for each model. This can be created by `plot(gy_logit, which = 6)` and
`plot(gy_loglog, which = 6)`, respectively.

In principle, the link function $g_2$ in the precision equation could
also influence the model fit. However, as the best-fitting model 
`gy_loglog` has a constant $\phi$, all links $g_2$ lead to 
equivalent estimates of $\phi$ and thus to equivalent fitted log-likelihoods.
However, the link function can have consequences in terms of the
inference about $\phi$ and in terms of convergence of the optimization.
Typically, a log-link leads to somewhat improved quadratic approximations
of the likelihood and less iterations in the optimization. For example,
refitting `gy_loglog` with $g_2(\cdot) = \log(\cdot)$ converges
more quickly:

```{r GasolineYield-loglog2}
gy_loglog2 <- update(gy_loglog, link.phi = "log")
summary(gy_loglog2)$iterations
```

with a lower number of iterations than for `gy_loglog`
which had `r summary(gy_loglog)$iterations` iterations.


#### Household food expenditures

One could conduct a similar analysis as above for the household
food expenditure data. However, as the response takes less extreme observations
than for the gasoline yield data, the choice of link function is less important.
In fact, refitting the model with various link functions shows no large
differences in the resulting log-likelihoods. These can be easily
extracted from fitted models using the `logLik()` function, e.g.,
`logLik(fe_beta2)`. Below we use a compact `sapply()` call to
obtain this for updated versions of `fe_beta2` with all available link
functions.

```{r FoodExpenditure-links}
sapply(c("logit", "probit", "cloglog", "cauchit", "loglog"),
  function(x) logLik(update(fe_beta2, link = x)))
```

Only the Cauchy link performs somewhat better than the logit link and might
hence deserve further investigation.



## Further replication exercises {#sec-replications}

In this section, further empirical illustrations of beta regressions
are provided. While the emphasis in the previous section was to present
how the various features of **betareg** can be used in pracice, we
focus more narrowly on replication of previously published research
articles below.

### Dyslexia and IQ predicting reading accuracy

```{r ReadingSkills-eda, echo=FALSE, results='hide'}
data("ReadingSkills", package = "betareg")
rs_accuracy <- format(round(with(ReadingSkills, tapply(accuracy, dyslexia, mean)), digits = 3))
```

We consider an application that analyzes reading accuracy data for nondyslexic
and dyslexic Australian children [@betareg:Smithson+Verkuilen:2006].  
The variable of interest is `accuracy` providing the scores on a test of 
reading accuracy taken by 44 children, which is predicted by the two regressors
`dyslexia` (a factor with sum contrasts separating a dyslexic and a control group)
and nonverbal intelligent quotient (`iq`, converted to $z$ scores), see
@fig-ReadingSkills for a visualization.
The sample includes 19 dyslexics and 25 controls who were recruited from primary 
schools in the Australian Capital Territory. The children's ages ranged from 
eight years five months to twelve years three months; mean reading accuracy was
`r rs_accuracy[2]` for dyslexic readers and `r rs_accuracy[1]` for controls. 

@betareg:Smithson+Verkuilen:2006 set out to investigate whether `dyslexia`
contributes to the explanation of `accuracy` even when corrected for `iq` score
(which is on average lower for dyslexics). Hence, they consider separate regressions
for the two groups fitted by the interaction of both regressors. To show that OLS
regression is no suitable tool in this situation, they first fit a linear regression
of the logit-transformed response:

```{r ReadingSkills-ols}
data("ReadingSkills", package = "betareg")
rs_ols <- lm(qlogis(accuracy) ~ dyslexia * iq, data = ReadingSkills)
coeftest(rs_ols)
```

The interaction effect does not appear to be significant, however this is a result
of the poor fit of the linear regression as will be shown below. @fig-ReadingSkills
clearly shows that the data are asymmetric and heteroskedastic (especially in the control
group). Hence, @betareg:Smithson+Verkuilen:2006 fit a beta regression model,
again with separate means for both groups, but they also allow the dispersion to depend on the
main effects of both variables.

```{r ReadingSkills-beta}
rs_beta <- betareg(accuracy ~ dyslexia * iq | dyslexia + iq,
  data = ReadingSkills, hessian = TRUE)
coeftest(rs_beta)
```

This shows that precision increases with `iq` and is lower for controls while
in the mean equation there is a significant interaction between `iq` and `dyslexia`.
As @fig-ReadingSkills illustrates, the beta regression fit does not differ
much from the OLS fit for the dyslexics group (with responses close to $0.5$) but fits
much better in the control group (with responses close to $1$).

The estimates above replicate those in Table 5 of @betareg:Smithson+Verkuilen:2006,
except for the signs of the coefficients of the dispersion submodel which they defined
in the opposite way. Note that their results have been obtained with numeric rather than
analytic standard errors hence `hessian = TRUE` is set above for replication. The results
are also confirmed by @betareg:Espinheira+Ferrari+Cribari-Neto:2008b, who have also
concluded that the dispersion is variable. As pointed out in @sec-vardisp, to 
formally test equidispersion against variable dispersion `lrtest(rs_beta, . ~ . | 1)`
(or the analogous `waldtest()`) can be used.

@betareg:Smithson+Verkuilen:2006 also consider two other psychometric applications of beta
regressions the data for which are also provided in the **betareg** package: see `?MockJurors`
and `?StressAnxiety`. Furthermore, `demo("SmithsonVerkuilen2006", package = "betareg")`
is a complete replication script with comments.

```{r ReadingSkills-visualization}
#| echo: false
#| fig-width: 6
#| fig.height: 5.5
#| out-width: 100%
#| label: fig-ReadingSkills
#| fig-cap: "Reading skills data from @betareg:Smithson+Verkuilen:2006 : Linearly transformed reading accuracy by IQ score and dyslexia status (control, blue vs. dyslexic, red). Fitted curves correspond to beta regression `rs_beta` (solid) and OLS regression with logit-transformed dependent variable `rs_ols` (dashed)."
cl1 <- hcl(c(260, 0), 90, 40)
cl2 <- hcl(c(260, 0), 10, 95)
plot(accuracy ~ iq, data = ReadingSkills, col = cl2[as.numeric(dyslexia)],
  main = "Reading skills data", xlab = "IQ score", ylab = "Reading accuracy",
  pch = c(19, 17)[as.numeric(dyslexia)], cex = 1.5)
points(accuracy ~ iq, data = ReadingSkills, cex = 1.5,
  pch = (1:2)[as.numeric(dyslexia)], col = cl1[as.numeric(dyslexia)])
nd <- data.frame(dyslexia = "no", iq = -30:30/10)
lines(nd$iq, predict(rs_beta, nd), col = cl1[1], lwd = 2)
lines(nd$iq, plogis(predict(rs_ols, nd)), col = cl1[1], lty = 2, lwd = 2)
nd <- data.frame(dyslexia = "yes", iq = -30:30/10)
lines(nd$iq, predict(rs_beta, nd), col = cl1[2], lwd = 2)
lines(nd$iq, plogis(predict(rs_ols, nd)), col = cl1[2], lty = 2, lwd = 2)
legend("topleft", c("control", "dyslexic", "betareg", "lm"),
  lty = c(NA, NA, 1:2), pch = c(19, 17, NA, NA), lwd = 2,
  col = c(cl2, 1, 1), bty = "n")
legend("topleft", c("control", "dyslexic", "betareg", "lm"),
  lty = c(NA, NA, 1:2), pch = c(1, 2, NA, NA),
  col = c(cl1, NA, NA), bty = "n")
```


### Structural change testing in beta regressions

As already illustrated in @sec-illustrations, `"betareg"` objects can be plugged into
various inference functions from other packages because they provide suitable
methods to standard generic functions (see @tbl-methods). Hence
`lrtest()` could be used for performing likelihood-ratio testing
inference and similarly
`coeftest()`, `waldtest()` from **lmtest** [@betareg:Zeileis+Hothorn:2002]
and `linear.hypothesis()` from **car** [@betareg:Fox+Weisberg:2019] can be
employed for carrying out different flavors of Wald tests.

In this section, we illustrate yet another generic inference approach implemented
in the **strucchange** package for structural change testing. This is concerned
with a different testing problem compared to the functions above: It assesses
whether the model parameters are stable throughout the entire sample or
whether they change over the observations $i = 1, \dots, n$. This is of particular
interest in time series applications where the regression coefficients $\beta$ and
$\gamma$ change at some unknown time in the sample period
[see @betareg:Zeileis:2006 for more details and references to the literature].

While originally
written for linear regression models [@betareg:Zeileis+Leisch+Hornik:2002],
**strucchange** was extended by @betareg:Zeileis:2006 to compute generalized
fluctuation tests for structural change in models that are based on suitable
estimating functions. The idea is to capture systematic deviations from parameter
stability by cumulative sums of the empirical estimating functions: If the
parameters are stable, the cumulative sum process should fluctuate randomly around
zero. However, if there is an abrupt shift in the parameters, the cumulative sums
will deviate clearly from zero and have a peak at around the time of the shift.
If the estimating functions can be extracted by an `estfun()` method (as for
`"betareg"` objects), models can simply be plugged into the `gefp()` function
for computing these cumulative sums (also known as generalized empirical
fluctuation processes). To illustrate this, we replicate the example from
Section 5.3 in @betareg:Zeileis:2006.

Two artificial data sets are considered: a series `y1` with a change in
the mean $\mu$, and a series `y2` with a change in the precision $\phi$.
Both simulated series start with the parameters $\mu = 0.3$ and $\phi = 4$
and for the first series $\mu$ changes to $0.5$ after 75% of the observations
while $\phi$ remains constant whereas for the second series $\phi$ changes to $8$
after 50% of the observations and $\mu$ remains constant.

```{r strucchange-data}
suppressWarnings(RNGversion("3.5.0"))
set.seed(123)
y1 <- c(rbeta(150, 0.3 * 4, 0.7 * 4), rbeta(50, 0.5 * 4, 0.5 * 4))
y2 <- c(rbeta(100, 0.3 * 4, 0.7 * 4), rbeta(100, 0.3 * 8, 0.7 * 8))
```

To capture instabilities in the parameters over "time" (i.e., the ordering of
the observations), the generalized empirical fluctuation processes can be derived
via

```{r strucchange-gefp}
library("strucchange")
y1_gefp <- gefp(y1 ~ 1, fit = betareg)
y2_gefp <- gefp(y2 ~ 1, fit = betareg)
```

and visualized by

```{r strucchange-plot1}
#| fig-width: 6.5
#| fig-height: 6
#| label: fig-strucchange1
#| fig-cap: "Structural change tests for artificial data `y1` with change in $\\mu$."
plot(y1_gefp, aggregate = FALSE)
```

```{r strucchange-plot2}
#| fig-width: 6.5
#| fig-height: 6
#| label: fig-strucchange2
#| fig-cap: "Structural change tests for artificial data `y2` with change in $\\phi$."
plot(y2_gefp, aggregate = FALSE)
```

The resulting @fig-strucchange1 and @fig-strucchange2 replicate Figure 4 from @betareg:Zeileis:2006
and show two 2-dimensional fluctuation processes: one for `y1` and one for `y2`.
Both fluctuation processes behave as expected: There is no excessive fluctuation of the 
process pertaining to the parameter that remained constant while there is a clear peak
at about the time of the change in the parameter with the shift. In both series
the structural change is significant due to the crossing of the red boundary that
corresponds to the 5% critical value. For further details see @betareg:Zeileis:2006.



## Summary {#sec-conclusion}

This paper addressed the R implementation of the class of beta regression models available in the
**betareg** package. We have presented the fixed and variable dispersion beta regression models, described
how one can model rates and proportions using **betareg** and presented several empirical examples
reproducing previously published results. Future
research and implementation shall focus on the situation where the data contain zeros and/or ones
[see e.g., @betareg:Kieschnick+McCullough:2003]. An additional line of research and
implementation is that of dynamic beta regression models, such as the class of 
$\beta$ARMA models
proposed by @betareg:Rocha+Cribari-Neto:2009.  


## Acknowledgments {.unnumbered}

FCN gratefully acknowledges financial support from CNPq/Brazil. 
Both authors are grateful to A.B. Simas and A.V. Rocha for their
work on the previous versions of the **betareg** package (up to
version 1.2). Furthermore, detailed and constructive feedback from
two anonymous reviewers, the associated editor, as well as from
B. Grün was very helpful for enhancing both software and manuscript.

## References {.unnumbered}
